---
title: "Reproducible Research: Peer Assessment 1"
author: "Xavier Merino"
date: "Friday, August 15, 2014"
output: html_document
---

## Loading and preprocessing the data
Check existence of Data folder and create it if it's not present
```{r}
if (!file.exists("./Data")){
  dir.create("./Data")
}
```

Unzipping and reading the data
```{r}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip","./Data/activity.zip")
unzip("./Data/activity.zip", exdir="./Data")
activity<-read.csv("./Data/activity.csv")
```
Setting the local language to english, to work with weekdays
```{r}
Sys.setlocale("LC_TIME", "English")
```
Storing the activity intervals as a variable for later use
```{r}
intervalsnum<-as.numeric(unique(activity[,"interval"]))
```
Converting the interval column into time format
```{r}
temp2 <- mapply(function(x, y) paste0(rep(x, y), collapse = ""), 0, 4 - nchar(activity[,"interval"]))
activity[,"interval"]<-paste0(temp2, activity[,"interval"])
activity[,"interval"]<-format(strptime(activity[,"interval"], format="%H%M"), format = "%H:%M")
```
Converting the date column into date format
```{r}
activity[,"date"]<-strftime(activity[,"date"],format="%Y-%m-%d")
```
Getting vectors with all the diferent dates and times 
```{r}
dates<-as.Date(unique(activity[,"date"]))
intervals<-unique(activity[,"interval"])
```

## What is mean total number of steps taken per day?

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
